Digital Analysis of Paintings

### 2014-05-09 paper.tex
Just conclusion to go :-)

This technique involves finding a function of $x$ and $y$ which steers a filter

% TODO I'm struggling to understand this one...

@@

% TODO Again, another complex one :-(

@@

>>>>

This technique involves finding a function of $x$ and $y$ which rotates a given
filter by a given angle, typically in terms of $\frac{\pi}{4}$ rotations, but
it can be more fine-grain than that.

A lot of this research stems from the idea that images are really
two-dimensional signals; therefore filters can also be thought of similarly.
And a function to rotate them will generally deal with shifting the signal
along an axis or transforming the signal so it has the effect of rotating the
signal.

This process can have high computational costs when used on a filter which is
not separable in terms of $x$ and $y$, so it is preferred if a filter can be
designed to be so, rather than take the computational cost.

These filters can be used for several applications, from analysing the local
orientation of gradients, which in turn can be used to build up systems which
detect items in images, such as \cite{dalal05histograms}; help to remove noise
in an image based on angles; detect contours within an image when applied to an
image transformed using an edge detection algorithm; and calculating the shape
from the shading of an image.

This paper presents an idea and its application but reads more like a textbook
reference than an experiment. As such there are no actual results detailed by
the paper other than to visualise some of the problems and solutions steerable
filters can be applied to.

Of course one could question the applicability of this research; but given the
wide-spread use of this research\footnote{Cited by at least 680 other IEEE
papers} this is very doubtable.

@@

Gabor filters are a sinusoidal wave within a Gaussian
envelope\cite{gabor1946theory}\cite{daugman1985uncertainty} which allows the
filter which matches signals with variable orientation, frequency and
bandwidth. They are useful within image processing as they can be steered in a
near limitless range of orientations by the change of a single parameter. The
size (frequency) of the areas can also be tuned with the modification of a
second parameter, making them very powerful.

These attributes make Gabor filters appropriate for the task of texture
discrimination. Texture discrimination is the task of labelling whether two
textures are the same or different. To the human vision model this is quite a
natural and simple take, but to a machine this is a complex task as a simple
equality check does not account for noise or changes in intensity.

Gabor filters are designed to model the way human vision system works as
closely as possible.

\cite{turner1986texture} describes a method of using Gabor filters for texture
discrimination and is done by passing Gabor filters at $\frac{\pi}{4}$
increments in orientation from $0$ to $\pi$ over overlapping regions of images
at two phases. The results of the passes of each filter are then made into a
phase insensitive measurement by taking the Euclidean distance of the two
measurements based on the result from the filter and the intensity of the
pixels.

Regions are then partitioned based on the sum of this measurement across a
square of pixels.

Although the results are very subjective; based on visualisations compared to
the human analysis, they do show some promising results, although in some
respects not entirely unexpected.

For example, Gabor filters ``see'' T-shaped structures closer to L-shaped
structures rather than T-shaped structures rotated by $\frac{\pi}{4}$. This
could be due to the low range of orientations the authors use in this paper.

This paper does show how useful Gabor filters can be in their application to
image processing; especially when considering the texture of images; however it
does also expose some of the flaws of a filter-only approach.

Although only using four orientations may appear to limit the applicability of
the filters, other research\cite{brown13can} also shows that higher
orientations does not always provide better results.

@@

Another avenue which can be taken is to use an artist's existing work to try
and classify new examples. \cite{brown13can} describes one attempt to do this
using a variety of techniques, from colour analysis to texture details.

To verify the results gained from this research the authors used a
leave-one-out cross-validation strategy on the 102 works for which a date was
know. This provided a correlation between actual year and classified year, as
well as the statistical significance of these results and the number of
paintings which were successfully classified within 15 years.

Analysis techniques provided results in the form of histograms which could then
be used to measure the distance between an example and the training set and
classified using $k$-nearest neighbour. This classification algorithm was used
because of its simplicity and naivety, which allowed it to only consider
distance in the feature space rather than any patterns in the underlying data.

Graphing the results of the classification a ``best'' value of $k = 7$ for
$k$-nearest neighbour was selected visually and used to gather formal results.
These results show correlations approaching $0.5$ for some of the more advanced
techniques; those examining the texture of the work.

This research also considered the idea of exemplar paintings; those paintings
which are most representative for a given year. Expert knowledge was used to
produce a list of \emph{Artistic Exemplars}, those most representative of the
artists style to an expert in the field. From analysis of a group of paintings
for each year a central point was computed, then the closest painting to this
central point was chosen as a \emph{Statistical Exemplar} for that year.

These exemplars were then used to classify other paintings by labelling the
painting with the nearest exemplar. Though these don't show as good correlation
results, it does allow other avenues of work to be created such as
investigating the differences between the \emph{Artistic} and
\emph{Statistical} exemplars.

This work does start tcentral point was computed, then the closest painting to
this central point was chosen as a \emph{Statistical Exemplar} for that year.

These exemplars were then used to classify other paintings by labelling the
painting with the nearest exemplar. Though these don't show as good correlation
results, it does allow other avenues of work to be created such as
investigating the differences between the \emph{Artistic} and
\emph{Statistical} exemplars.

This work is obviously in its infancy but does show some promising results. The
methodology is well thought through.

The authors do discuss the use of $k$-nearest neighbour and whether it should
be compared to other classifiers. An \gls{SVM} seems particularly applicable as
it would help draw out some of the underlying patterns in the artists work.

The lack of high-resolution images with non-standard lighting has hindered this
research somewhat; although part of the point of the research is that the
system should be robust enough to handle simple catalogue images, it does
appear to damage the performance.

It is also very specific to the artist Kyffin Williams, as his work was highly
textural, which might explain why features including texture performed better.
More work could be done to explore other artists and see how they compare to
Williams.

As the work uses histograms, it might have provided better results to combine
the results of the analysis techniques to gather a larger histogram. This would
have to be normalised or weighted based on the performance of the techniques to
gather meaningful analysis, but is another avenue which hasn't been explored
sufficiently.
